---
title: "Course Poject 2: The Impact of Various Weather Events on US Population Health and Economy"
output: html_document
---
# 1. Synopsis

The basic goal of this assignment is to explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database from 1950 to November 2011 and address the following questions:

* Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?

* Across the United States, which types of events have the greatest economic consequences?

The result shows that `TORNADO` is the major event type which greatly affects the population headlth (most number of fatalities and injuries). Meanwhile, `FLOOD` is the event type that greatly affects US economy particularly on Properties and Crop.

# 2. Data Processing

## 2.1 Data Downloading and Reading
Download the data (compressed in bz2 format) from the course website and read the file using `read.csv()` to load the data into R (Note: `read.csv()` can read the .csv file even if it's compressed in .bz2)
```{r}
if (!file.exists("repdata_data_StormData.csv.bz2")) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","repdata_data_StormData.csv.bz2")
}

if (!exists("storm")) {
  storm <- read.csv("repdata_data_StormData.csv.bz2")
}
```

## 2.2 Data Exploration
```{r}
str(storm)
```
The `str()` function is a very handy tools to explore the data. There are 902297 observations with 37 variables which makes the data difficult to view in one whole screen. To summarize and simplify our dataset, a subset dataset was created `st` containing 8 variables only ("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP","CROPDMG", and "CROPDMGEXP"). 

There are 985 different event types on the `EVTYPE` variable. Looking at the `EVTYPE` data, alot of typo errors where observe like TORNDAO instead of TORNADO. Other entries have the same meaning but were recorded differently like FLASH FLOOD and FLOOD/FLASH. Some were written in uppercase and others are in lowercase but only refers to one event type.

The following code performs a text processing, grouping the same weather events as precisely as possible. After processing, there are 262 different event types left, much less than the original number of types, 985. 

## 2.3 Subset Needed Variables
```{r}
# Selecting the needed variables (columns)
st <- storm[, c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP")]
```

## 2.4 Cleaning the EVTYPE variable
```{r}
# Convert lower cases event type to upper class to reduce levels
st$EVTYPE <- factor(toupper(st$EVTYPE))
# EVTYPE levels decreased from 985 to 898 (Use str() function to check)

# Fixing / triming event types that are redundant just like with spaces
trim <- function(x) {gsub("(^[[:space:]]+|[[:space:]]+$)", "", x)}
st$EVTYPE <- factor(trim(st$EVTYPE))
# EVTYPE levels decreased from 898 to 890

# Combining mispelled event types to correct ones using grepl to reduce event types/levels
st[grepl("BLIZZARD", st$EVTYPE), ]$EVTYPE <- "BLIZZARD"
st[grepl("AVALANCE", st$EVTYPE), ]$EVTYPE <- "AVALANCHE"
st[grepl("BRUSH FIRE", st$EVTYPE), ]$EVTYPE <- "BRUSH FIRE"
st[grepl("BLOW-OUT TIDE", st$EVTYPE), ]$EVTYPE <- "BLOW-OUT TIDE"
st[grepl("DRY", st$EVTYPE), ]$EVTYPE <- "DRY"
st[grepl("DROUGHT", st$EVTYPE), ]$EVTYPE <- "DROUGHT"
st[grepl("COASTAL FLOOD", st$EVTYPE), ]$EVTYPE <- "COASTAL FLOOD"
st[(grepl("COLD", st$EVTYPE) & !(grepl("EXTREME COLD", st$EVTYPE)) & !(grepl("EXTREME/RECORD COLD", st$EVTYPE))), ]$EVTYPE <- "COLD"
st[grepl("EXTREME/RECORD COLD", st$EVTYPE), ]$EVTYPE <- "EXTREME COLD"
st[grepl("DUST STORM", st$EVTYPE), ]$EVTYPE <- "DUSTSTORM"
st[grepl("DUST DEVEL", st$EVTYPE), ]$EVTYPE <- "DUST DEVIL"
st[grepl("FLASH FLOOD", st$EVTYPE), ]$EVTYPE <- "FLASH FLOOD"
st[grepl("FLOOD FLASH", st$EVTYPE), ]$EVTYPE <- "FLASH FLOOD"
st[grepl("FLOOD/FLASH", st$EVTYPE), ]$EVTYPE <- "FLASH FLOOD"
st[(grepl("FLOOD", st$EVTYPE) & !(grepl("FLASH FLOOD", st$EVTYPE)) & !(grepl("LAKESHORE FLOOD", st$EVTYPE)) & !(grepl("COASTAL FLOOD", st$EVTYPE))), ]$EVTYPE <- "FLOOD"
st[grepl("FUNNEL", st$EVTYPE), ]$EVTYPE <- "FUNNEL CLOUD"
st[grepl("FROST", st$EVTYPE), ]$EVTYPE <- "FROST/FREEZE"
st[grepl("GLAZE", st$EVTYPE), ]$EVTYPE <- "GLAZE"
st[grepl("HAIL", st$EVTYPE) & !(grepl("MARINE HAIL", st$EVTYPE)), ]$EVTYPE <- "HAIL"
st[grepl("HEAT", st$EVTYPE) & !(grepl("EXCESSIVE HEAT", st$EVTYPE)), ]$EVTYPE <- "HEAT"
st[grepl("HEAVY RAIN", st$EVTYPE), ]$EVTYPE <- "HEAVY RAIN"
st[grepl("HEAVY PRECIPATATION", st$EVTYPE), ]$EVTYPE <- "HEAVY PRECIPITATION"
st[grepl("HIGH  SWELLS", st$EVTYPE), ]$EVTYPE <- "HIGH SWELLS"
st[grepl("HEAVY SHOWERS", st$EVTYPE), ]$EVTYPE <- "HEAVY SHOWER"
st[grepl("HVY RAIN", st$EVTYPE), ]$EVTYPE <- "HEAVY RAIN"
st[grepl("HEAVY SNOW", st$EVTYPE), ]$EVTYPE <- "HEAVY SNOW"
st[grepl("HIGH SURF", st$EVTYPE), ]$EVTYPE <- "HIGH SURF"
st[grepl("HIGH WIND", st$EVTYPE) & !(grepl("MARINE HIGH WIND", st$EVTYPE)), ]$EVTYPE <- "HIGH WIND"
st[grepl("HURRICANE", st$EVTYPE), ]$EVTYPE <- "HURRICANE"
st[grepl("HYPOTHERMIA/EXPOSURE", st$EVTYPE), ]$EVTYPE <- "HYPOTHERMIA"
st[grepl("ICE ON ROAD", st$EVTYPE), ]$EVTYPE <- "ICE ROADS"
st[grepl("ICY ROADS", st$EVTYPE), ]$EVTYPE <- "ICE ROADS"
st[grepl("LANDSLIDES", st$EVTYPE), ]$EVTYPE <- "LANDSLIDE"
st[grepl("LAKE EFFECT SNOW", st$EVTYPE), ]$EVTYPE <- "LAKE-EFFECT SNOW"
st[grepl("LIGHTNING", st$EVTYPE), ]$EVTYPE <- "LIGHTNING"
st[grepl("LIGHTING", st$EVTYPE), ]$EVTYPE <- "LIGHTNING"
st[grepl("LIGNTNING", st$EVTYPE), ]$EVTYPE <- "LIGHTNING"
st[grepl("LOW TEMPERATURE RECORD", st$EVTYPE), ]$EVTYPE <- "LOW TEMPERATURE"
st[grepl("MUD SLIDE", st$EVTYPE), ]$EVTYPE <- "MUDSLIDE"
st[grepl("MUDSLIDE/LANDSLIDE", st$EVTYPE), ]$EVTYPE <- "MUDSLIDE"
st[grepl("MUD/ROCK SLIDE", st$EVTYPE), ]$EVTYPE <- "MUDSLIDE"
# st[grepl("RAIN (HEAVY)", st$EVTYPE), ]$EVTYPE <- "HEAVY RAIN"
st$EVTYPE[289860] <- "HEAVY RAIN"
st[grepl("RIP CURRENT", st$EVTYPE), ]$EVTYPE <- "RIP CURRENT"
st[grepl("RECORD/EXCESSIVE HEAT", st$EVTYPE), ]$EVTYPE <- "EXCESSIVE HEAT"
st[grepl("SMALL STREAM AND", st$EVTYPE), ]$EVTYPE <- "SMALL STREAM"
st[grepl("SML STREAM FLD", st$EVTYPE), ]$EVTYPE <- "SMALL STREAM"
st[grepl("TORRENTIAL RAINFALL", st$EVTYPE), ]$EVTYPE <- "TORRENTIAL RAIN"
st[grepl("THUNDERSTORM WIND", st$EVTYPE) & !(grepl("MARINE THUNDERSTORM WIND", st$EVTYPE)), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUDERSTORM WINDS", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNDEERSTORM WINDS", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNDERESTORM WINDS", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNDERSTORM", st$EVTYPE) & !(grepl("MARINE THUNDERSTORM WIND", st$EVTYPE)), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("TUNDERSTORM", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNDERSTROM", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNDERTSORM", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNDERTORM", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNDESTORM", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("THUNERSTORM", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("SNOW", st$EVTYPE), ]$EVTYPE <- "SNOW"
st[grepl("SLEET", st$EVTYPE), ]$EVTYPE <- "SLEET"
st[grepl("STORM SURGE", st$EVTYPE), ]$EVTYPE <- "STORM SURGE"
st[grepl("TYPHOON", st$EVTYPE), ]$EVTYPE <- "HURRICANE"
st[grepl("TSTM", st$EVTYPE), ]$EVTYPE <- "THUNDERSTORM WIND"
st[grepl("TORNADO", st$EVTYPE), ]$EVTYPE <- "TORNADO"
st[grepl("TORNDAO", st$EVTYPE), ]$EVTYPE <- "TORNADO"
st[grepl("UNSEASONABLY WARM YEAR", st$EVTYPE), ]$EVTYPE <- "UNSEASONABLY WARM"
st[grepl("UNUSUAL/RECORD WARMTH", st$EVTYPE), ]$EVTYPE <- "UNUSUAL WARMTH"
st[grepl("UNUSUALLY WARM", st$EVTYPE), ]$EVTYPE <- "UNUSUAL WARMTH"
st[grepl("URBAN", st$EVTYPE), ]$EVTYPE <- "URBAN SMALL"
st[grepl("UNSEASONABLY WARM & WET", st$EVTYPE), ]$EVTYPE <- "UNSEASONABLY WARM/WET"
st[grepl("WET MICOBURST", st$EVTYPE), ]$EVTYPE <- "WET MICROBURST"
st[grepl("VOLCANIC ASH", st$EVTYPE), ]$EVTYPE <- "VOLCANIC ASH"
st[grepl("WATERSPOUT", st$EVTYPE), ]$EVTYPE <- "WATERSPOUT"
st[grepl("WATER SPOUT", st$EVTYPE), ]$EVTYPE <- "WATERSPOUT"
st[grepl("WAYTERSPOUT", st$EVTYPE), ]$EVTYPE <- "WATERSPOUT"
st[grepl("TROPICAL STORM", st$EVTYPE), ]$EVTYPE <- "TROPICAL STORM"
st[grepl("WILDFIRE", st$EVTYPE), ]$EVTYPE <- "WILDFIRE"
st[grepl("WILD FIRES", st$EVTYPE), ]$EVTYPE <- "WILDFIRE"
st[grepl("WILD/FOREST FIRE", st$EVTYPE), ]$EVTYPE <- "WILDFIRE"
st[grepl("WINTER STORM", st$EVTYPE), ]$EVTYPE <- "WINTER STORM"
st[grepl("WINTER WEATHER", st$EVTYPE), ]$EVTYPE <- "WINTER WEATHER"
st[(grepl("WIND", st$EVTYPE) & !(grepl("MARINE THUNDERSTORM WIND", st$EVTYPE)) & !(grepl("THUNDERSTORM WIND", st$EVTYPE))), ]$EVTYPE <- "WIND"
st[grepl("WND", st$EVTYPE), ]$EVTYPE <- "WIND"
st[grepl("WINTERY MIX", st$EVTYPE), ]$EVTYPE <- "WINTER MIX"
st[grepl("WINTRY MIX", st$EVTYPE), ]$EVTYPE <- "WINTER MIX"
st$EVTYPE <- trim(st$EVTYPE)
# EVTYPE levels decreased from 890 to 262 (use unique(st$EVTYPE) to check # of event types)
```

# 3. Results

## 3.1 Health consequences of severe weather

### A. Fatalities
Looking at the top causes of fatalities :
```{r}
# Aggregate event types by number of fatalities
fatal <- aggregate(FATALITIES ~ EVTYPE, data = st, sum)
# Determine the frequency of the event type occcurence
freq <- as.matrix(table(st$EVTYPE))
# Combine the no. of fatalities, frequency and fatality rate
fatal <- cbind(fatal, freq, rate = fatal$FATALITIES / freq)
# Arrange in descending order based on the number of fatalities
fatal_order <- fatal[order(fatal$FATALITIES, decreasing = TRUE), ]
# Look at the top 10 event types that caused most fatalities
head(fatal_order, n = 10)
```

The `TORNADO`, `(EXCESSIVE) HEAT` and `FLASH FLOOD` caused the most fatalities. But this may be due to high occurrences of those events as seen on the figure below.

```{r, message=FALSE}
library(ggplot2)
theme <- theme(axis.text.x = element_text(colour = "black", size = 12, angle = 90), 
          axis.text.y = element_text(colour = "grey20", size = 12),
          axis.title.x = element_text(size = 12, hjust = .5),
          axis.title.y = element_text(size = 12, hjust = .5, vjust = 1),
          title = element_text(size = 14, hjust = .5)) +
    theme(legend.position = "none")
g <- ggplot(fatal_order[1:10, ], aes(reorder(EVTYPE, -FATALITIES), FATALITIES))
g <- g + geom_bar(stat = "identity", aes(fill = FATALITIES)) + 
    xlab("Event Type") + ylab("Number of Fatalities") + 
    ggtitle('Top 10 Event Types with Most Fatalities') + theme
print(g)
```

```{r}
# Arrange in descending order based on the number of fatality rate
fatal_rate_order <- fatal[order(fatal$rate, decreasing = TRUE), ]
# Look at the top 10 event types with highest fatality rate
head(fatal_rate_order, n = 10)
# Determine the number of types of events causing fatalities
(sum(fatal$FATALITIES >= 1))
```

If looking on the fatality rate (number of fatalities / number of event occured), the `MARINE MISHAP`, `ROUGH SEAS`, and `TSUNAMI`have the highest fatality rate since these events frequently occur but bring more casualties compared to other events.

### B. Injuries
Looking at the top causes of injuries :

```{r, fig.height=5, fig.width=10}
# Aggregate event types by number of injuries
injury <- aggregate(INJURIES ~ EVTYPE, data = st, sum)
# Arrange in descending order based on the number of injuries
injury_order <- injury[order(injury$INJURIES, decreasing = TRUE),]
# Look at the top 10 event types that caused most injuries
head(injury_order, n = 10)
```

`TORNADO` also topped the causes of injuries.

```{r}
g_in <- ggplot(injury_order[1:10, ], aes(reorder(EVTYPE, -INJURIES), INJURIES))
g_in <- g_in + geom_bar(stat = "identity", aes(fill = INJURIES)) + 
    xlab("Event Type") + ylab("Number of Injuries") + 
    ggtitle('Top 10 Event Types with Most Injuries') + theme
print(g_in)
```

```{r}
# Determine the frequency of the event type occcurence
freq <- as.matrix(table(st$EVTYPE))
# Combine the no. of injuries, frequency and fatality rate
injury <- cbind(injury, freq, rate = injury$INJURIES / freq)
# Arrange in descending order based on the number of fatalities
injury_rate_order <- injury[order(injury$rate, decreasing = TRUE), ]
# Look at the top 10 event types with highest fatality rate
head(injury_rate_order, n = 10)
# Determine the number of types of events causing injuries
(sum(injury$INJURIES>=1)) 
```
Looking the result for the highest injury rate, `TORNADO` also tops the list.


## 3.2 Economic Consequences

### A. Property and Crop Damages

To calculate the damage expenses correctly, First, convert the recorded damage numbers into expenses in dollar unit.

```{r, fig.height=5, fig.width=12}
unique(st$PROPDMGEXP)
unique(st$CROPDMGEXP)
idx <- c("", "+", "-", "?", 0:9, "h", "H", "k", "K", "m", "M", "b", "B");
digit <- c(rep(0,4), 0:9, 2, 2, 3, 3, 6, 6, 9, 9)
multi <- data.frame(idx, digit)
st$damage.p <- st$PROPDMG * 10 ^ multi[match(st$PROPDMGEXP, multi$idx), 2]
st$damage.c <- st$CROPDMG * 10 ^ multi[match(st$CROPDMGEXP, multi$idx), 2]
st$damage <- st$damage.p + st$damage.c
dmg.p <- aggregate(damage.p ~ EVTYPE, data = st, sum)
dmg.p_order <- dmg.p[order(dmg.p$damage.p, decreasing = TRUE), ]
dmg.c <- aggregate(damage.c ~ EVTYPE, data = st, sum)
dmg.c_order <- dmg.c[order(dmg.c$damage.c, decreasing = TRUE), ]
dmg <- aggregate(damage ~ EVTYPE, data = st, sum)
dmg_order <- dmg[order(dmg$damage, decreasing = TRUE), ]
```

The figure below shows that `FLOOD` and `HURRICANE` are the top causes of property damages, while `DROUGHT` and `FLOOD` are the top causes of damages on crops. Overall, the `FLOOD` still caused the most economic loss.

```{r, include=FALSE}
# Create multi panel plot
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
    require(grid)
    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)
    numPlots = length(plots)
    
    # If layout is NULL, then use 'cols' to determine layout
    if (is.null(layout)) {
        # Make the panel
        # ncol: Number of columns of plots
        # nrow: Number of rows needed, calculated from # of cols
        layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                         ncol = cols, nrow = ceiling(numPlots/cols))
    }
    
    if (numPlots==1) {
        print(plots[[1]])
        
    } else {
        # Set up the page
        grid.newpage()
        pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
        
        # Make each plot, in the correct location
        for (i in 1:numPlots) {
            # Get the i,j matrix positions of the regions that contain this subplot
            matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
            print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                            layout.pos.col = matchidx$col))
        }
    }
}
```

```{r}
g.p <- ggplot(dmg.p_order[1:10, ], aes(reorder(EVTYPE, -damage.p), damage.p))
g.p <- g.p + geom_bar(stat = "identity", aes(fill = damage.p)) + 
    xlab("Event Type") + ylab("Property Damage ($)") + 
    ggtitle('Top 10 Property Damage') + theme
g.c <- ggplot(dmg.c_order[1:10, ], aes(reorder(EVTYPE, -damage.c), damage.c))
g.c <- g.c + geom_bar(stat = "identity", aes(fill = damage.c)) + 
    xlab("Event Type") + ylab("Crop Damage ($)") + 
    ggtitle('Top 10 Crop Damage') + theme
g.t <- ggplot(dmg_order[1:10, ], aes(reorder(EVTYPE, -damage), damage))
g.t <- g.t + geom_bar(stat = "identity", aes(fill = damage)) + 
    xlab("Event Type") + ylab("Total Damage ($)") + 
    ggtitle('Top 10 Total Damage') + theme
multiplot(g.p, g.c, g.t, cols = 3)
```

# 4. Summary

This project had successfully answered the following questions:

* Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health? The `TORNADO` is the major event type which greatly affects the population headlth (most number of fatalities and injuries).

* Across the United States, which types of events have the greatest economic consequences? 
 `FLOOD` is the event type that greatly affects US economy particularly on Properties and Crop.

